---
title: "Lite model: rice growth under flooded vs rainfed conditions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lite model: rice growth under flooded vs rainfed conditions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
library(spacsysr)
```

## Overview

This vignette demonstrates the lite model on rice. Rice is grown here in a
warm summer season with three scenarios: continuous flooding with fertiliser,
rainfed (no irrigation) with fertiliser, and flooding without fertiliser.
Flooding is represented by daily irrigation replacing evapotranspiration
demand (`irrig` column); there is no standing-water layer and no CH4
coupling in the lite model, so the focus is on water/N stress and the
yield vs. N2O trade-off of flooded soils.

## Minimal inputs

```{r}
set.seed(21)
n <- 150
weather <- data.frame(
  date   = seq(as.Date("2026-06-01"), by = "day", length.out = n),
  tmax   = 31 + 3 * sin(2 * pi * (1:n) / n) + rnorm(n, 0, 1.5),
  tmin   = 23 + 2 * sin(2 * pi * (1:n) / n) + rnorm(n, 0, 1.2),
  precip = pmax(0, rnorm(n, 2.5, 6))  # sub-humid: rainfed faces dry spells
)
w <- weather_complete(weather, lat_deg = 30, t_base = 8)

# Flooded: irrigation replaces ET demand each day, keeping soil near FC
w_flood <- w
w_flood$irrig <- pmax(0, w_flood$pet - w_flood$precip)

soil <- soil_init(depth_mm = c(200, 300, 500),
                  soc = c(1000, 600, 300),   # modest fertility: no-N shows stress
                  nh4_init = 0.3, no3_init = 0.5)
crop <- crop_default_params("rice")
```

## Scenarios

```{r}
fert <- data.frame(date = as.Date(c("2026-06-08", "2026-07-10", "2026-08-15")),
                   nh4_add = c(6, 4, 3), no3_add = c(0, 0, 0))

run <- function(w_in, fert_in) {
  spacsys_lite_run(w_in, soil, crop, n_inputs = fert_in, lat_deg = 30)
}
scenarios <- rbind(
  cbind(run(w_flood, fert),   scenario = "flooded + N"),
  cbind(run(w, fert),         scenario = "rainfed + N"),
  cbind(run(w_flood, NULL),   scenario = "flooded, no N")
)
```

## Canopy and yield

```{r}
cols <- c("steelblue", "firebrick", "darkgreen")
matplot(scenarios$date[1:n],
        sapply(split(scenarios$lai, scenarios$scenario), identity),
        type = "l", lty = 1, col = cols, xlab = "Date", ylab = "LAI",
        main = "Rice canopy development")
legend("topleft", legend = levels(factor(scenarios$scenario)),
       col = cols, lty = 1, bty = "n")

tapply(scenarios$w_grain, scenarios$scenario,
       function(x) round(tail(x, 1) / 100, 2))  # t/ha
```

## Stress, N uptake and N2O

```{r}
agg <- aggregate(cbind(f_w, f_n) ~ scenario, data = scenarios,
                 FUN = function(x) round(sum(x < 0.9), 0))
names(agg)[2:3] <- c("water_stress_days", "N_stress_days")
agg

tapply(scenarios$n_uptake, scenarios$scenario, sum)  # g N m-2, seasonal
round(tapply(scenarios$n2o, scenarios$scenario, sum) * 1000, 1)  # mg N m-2
```

Flooding removes water stress and gives the highest yield, but the wet soil
also drives more denitrification and hence higher N2O -- the classic
paddy yield-vs-emissions trade-off. Without fertiliser, growth is N-limited
despite ample water. Caveats: the lite model has no ponded-water layer, no
methane module and no iron-plaque/redox chemistry; treat absolute N2O as
illustrative.
